---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library(dplyr)
activity <- read.csv("activity.csv")
activity <- mutate(activity, datetime = as.POSIXct(date) + interval*60)
```

## What is mean total number of steps taken per day?

```{r}
activityNoNA <- activity[!is.na(activity$steps),]
stepsPerDay <- aggregate(steps ~ date, activityNoNA, sum)
hist(stepsPerDay$steps, main = "Histogram of the number of steps taken each day", xlab="Steps")
stepMean <- mean(stepsPerDay$steps)
stepMedian <- median(stepsPerDay$steps)
```
The mean total number of steps per day is `r stepMean`

The median total number of steps per day is `r stepMedian`

## What is the average daily activity pattern?

```{r}
stepsPerInterval <- aggregate(steps ~ interval, activityNoNA, mean)
plot(stepsPerInterval$interval,stepsPerInterval$steps, type="l", xlab="Time Interval",ylab="Average number of steps taken")
maxStepsIndex <- which.max(stepsPerInterval$steps)
interval <- stepsPerInterval$interval[maxStepsIndex]
```
The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r interval`

## Imputing missing values

```{r}
numOfNA <- sum(is.na(activity$steps))
```
The total number of missing values is `r numOfNA`

```{r}
replacedNA <- activity
# find the indicies of the NA values
ind <- which( is.na( replacedNA$steps ) )
# replace them with the mean for the corresponding interval
for(i in ind){
  replacedNA$steps[i] <- stepsPerInterval$steps[which(stepsPerInterval$interval == replacedNA$interval[i])]
}

stepsPerDay <- aggregate(steps ~ date, replacedNA, sum)
hist(stepsPerDay$steps, main = "Histogram of the number of steps taken each day", xlab="Steps")
stepMeanReplaced <- mean(stepsPerDay$steps)
stepMedianReplaced <- median(stepsPerDay$steps)
```
The mean total number of steps per day for the new data set is `r stepMeanReplaced`

The median total number of steps per day for the new data set is `r stepMedianReplaced`

Replacing the missing values with the interval mean has left the mean unchanged, but has shifted the median by `r stepMedianReplaced - stepMedian`

Imputting the missing data has increased the number of days which have a total number of steps in the range 10000 - 15000.


## Are there differences in activity patterns between weekdays and weekends?

```{r}
replacedNA$weekfactor <- ifelse(weekdays(replacedNA$datetime) %in% c("Saturday", "Sunday"), "weekend", "weekday")
stepsPerIntervalWeekday <- aggregate(steps ~ interval + weekfactor, replacedNA, mean)
par(mfrow=c(2,1))
weekendIndex <- which(stepsPerIntervalWeekday$weekfactor == "weekend")
weekdayIndex <- which(stepsPerIntervalWeekday$weekfactor == "weekday")
plot(stepsPerIntervalWeekday$interval[weekendIndex],stepsPerIntervalWeekday$steps[weekendIndex], type="l", xlab="Time Interval",ylab="Average number of steps taken", main="Weekend")
plot(stepsPerIntervalWeekday$interval[weekdayIndex],stepsPerIntervalWeekday$steps[weekdayIndex], type="l", xlab="Time Interval",ylab="Average number of steps taken", main="Weekday")
```

